---
title: "Data Wrangling and Manipulation"
subtitle: "Health Dataset Example – The Graph Courses"
author: "Isaac Tetteh Martey"
date: "March 2, 2026"
output:
  prettydoc::html_pretty:
    theme: cosmo
    highlight: github
    fig_caption: true
    df_print: paged         # nice interactive tables for data frames
---

```{r setup, include = FALSE}

# Global defaults — hide code + clean output for portfolio style
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align  = "center",
  out.width  = "80%",
  fig.asp    = 0.618,
  dpi     = 140
)

# Load packages once, quietly, in setup chunk
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
pacman::p_load(
  tidyverse,     # ggplot2, dplyr, readr, tidyr, etc.
  janitor,       # clean_names(), etc.
  gt,            # beautiful tables (better than kable for HTML)
  reactable,     # interactive tables — great for portfolio
  here           # optional — use if you switch to local files later
)
```

# Data Wrangling

This document entails initial data ingestion, cleaning, and structural transformation of the some health dataset from the <https://www.thegraphcourses.org>

The direct link from the github account of **thegraphcourses** were used so as to get the most updated clean dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}
yaounde_covid <- read_csv("https://raw.githubusercontent.com/the-graph-courses/idar_book/refs/heads/main/data/yaounde_data2.csv", show_col_types = FALSE) %>% clean_names()

```

```{r echo=TRUE, message =FALSE, warning=FALSE}
sacro_data <- read_csv("https://raw.githubusercontent.com/the-graph-courses/idar_book/refs/heads/main/data/sarcopenia_elderly.csv" , show_col_types = FALSE) %>% clean_names() 
```

```{r echo=TRUE, message =FALSE, warning=FALSE}
influenza <- read_csv("https://raw.githubusercontent.com/the-graph-courses/idar_book/refs/heads/main/data/flu_h7n9_china_2013.csv")
```

## Focusing on User Information for the Yaounde Dataset

The Yaoundé COVID-19 dataset contains 37 variables (demographics, symptoms, serology, behaviors, etc.).

For the initial exploration and most of the analysis in this portfolio, we focus on the **core participant characteristics** — from `age` to `occupation`.

This reduces the working dataset from 37 columns to a more manageable subset.

```{r}
# Select relevant demographic / user-info columns
yao <- yaounde_covid %>% 
  select(age:occupation)

yao %>% 
  slice_sample(n = 10) %>%           # random sample looks more natural than head()
  gt::gt() %>% 
  tab_header(
    title = "Sample of Core Participant Information",
    subtitle = "Yaoundé COVID-19 dataset – selected variables (random 10 rows)"
  ) %>% 
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  )
```

In the table above only 5 columns were selected using the **select(age:occupation)** but in the table below the **select(!c(highest_education:consultation))** was used to select all other columns than the ones in the concatenate function as th **!** represents NOT.

```{r}
yaounde_covid %>% 
  select(!c(highest_education:consultation)) %>% 
  slice_sample(n = 10) %>%           # random sample looks more natural than head()
  gt::gt() %>% 
  tab_header(
    title = "Sample of Core Participant Information",
  ) %>% 
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  )

```

## Selecting Binary / Indicator Variables

Many variables in the Yaoundé COVID-19 dataset are binary indicators (yes/no questions) whose columns start with `is_`. These typically represent:

-   Presence of symptoms (e.g. **is_fever** , **is_cough**)
-   Behaviors or exposures (e.g. **is_smoker**, **is_pregnant**)

I extract the participant **id** along with all the **is\_** columns to focus on this group of variables for later analysis (e.g. prevalence of symptoms, treatment patterns).

```{r}
yaounde_covid %>% 
  select(id, starts_with("is_")) %>% 
reactable(
  defaultPageSize = 12,
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  bordered = TRUE,
  columns = list(
    id = colDef(name = "Participant ID", minWidth = 110)
  ),
  defaultColDef = colDef(minWidth = 95, align = "center")
)
```


## Reordering Columns: Bringing Outcome Variables First
```{r echo=TRUE, eval=FALSE}
yaounde_covid %>% 
  select(ends_with("result"), everything()) 
```
- It reorders the columns so that:

- All columns whose names end with "result" come first (likely test/serology outcomes like igg_result, igm_result, etc.)
Then all the other columns follow in their original order

```{r}
yaounde_covid %>% 
  select(ends_with("result"), everything()) %>%
  select(1:8) %>%                         # show only first 8 columns (the important ones + a few others)
  slice_sample(n = 10) %>%                # random sample – more representative
  gt::gt() %>%
  tab_header(
    title = "Reordered Dataset Preview",
    subtitle = "Outcome/result columns moved to the front (random 10 participants)"
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  )
```

## Subsetting by Sex – Focusing on Male Participants

To explore potential sex-based differences (common in COVID-19 research — e.g. severity, symptoms), we filtered only male participants from the core user information dataset **yao**.

```{r}
yao %>% 
  filter(sex == "Male") %>%
  slice_sample(n = 12) %>%
  gt::gt() %>%
  tab_header(
    title = "Random Sample of Male Participants",
    subtitle = "Males in the Yaoundé COVID-19 dataset"
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  )
```

## Neighborhood Subsetting – Excluding / Focusing on Specific Areas

The Yaoundé dataset includes a **neighborhood** variable representing participants' residential areas.  

Two common analysis patterns are:

- **Excluding** certain neighborhoods 
- **Focusing only on** specific neighborhoods (e.g. to compare Tsinga and Messa against the rest of the sample)

I demonstrate both approaches below using the core dataset **yaounde_covid**.

```{r}
yaounde_covid %>% 
  filter(!(neighborhood %in% c("Tsinga", "Messa"))) %>% 
  count(neighborhood, name = "n_participants") %>%
  arrange(desc(n_participants)) %>%
  gt::gt() %>%
  tab_header(
    title = "Most Common Neighborhoods (Excluding Tsinga and Messa)",
  ) %>%
  tab_options(table.font.size = px(13)) %>%
  cols_label(
    neighborhood   = "Neighborhood",
    n_participants = "Number of Participants"
  ) %>% 
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_column_labels()
)
```


```{r}
yaounde_covid %>% 
  filter(neighborhood %in% c("Tsinga", "Messa")) %>%
  count(neighborhood, name = "n_participants") %>%
  gt::gt() %>%
  tab_header(
    title = "Participants from Tsinga and Messa Only",
  ) %>%
  tab_options(table.font.size = px(13)) %>%
  cols_label(
    neighborhood   = "Neighborhood",
    n_participants = "Number of Participants"
  ) %>%
  fmt_number(columns = n_participants, decimals = 0) %>% 
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_column_labels()
)
```

## Handling Missing Values in Key Variables – Pregnancy Status

The variable `is_pregnant` is a binary indicator that is only relevant/applicable for female participants.  

Many datasets record pregnancy status as `NA` for males, children, or when the question was skipped/not applicable.  

I create a subset that keeps only rows where pregnancy status is explicitly recorded (i.e. non-missing), which is useful before any pregnancy-related analysis.


```{r filter-known-pregnancy}
# Keep only rows where is_pregnant is not missing
known_pregnancy <- yaounde_covid %>%
  filter(!is.na(is_pregnant))

# Quick summary table – shows what we kept
known_pregnancy %>%
  count(is_pregnant, name = "n_participants") %>%
  mutate(
    percent     = round(n_participants / nrow(known_pregnancy) * 100, 1),
status = case_when(
    is_pregnant %in% c("Yes", "yes") ~ "Pregnant",
    is_pregnant %in% c("No",  "no") ~ "Not pregnant",
    TRUE ~ "Other / Invalid"
  )
) %>% 
  select(status, n_participants, percent) %>%
  gt::gt() %>%
  tab_header(
    title = "Pregnancy Status – Among Participants with Known Status",
    subtitle = paste("n =", nrow(known_pregnancy), "participants with non-missing is_pregnant")
  ) %>%
  tab_options(table.font.size = px(13)) %>%
  cols_label(
    status         = "Pregnancy Status",
    n_participants = "Number of Participants",
    percent        = "Percentage"
  ) %>%
  fmt_number(columns = c(n_participants, percent), decimals = 1) %>%
  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) 
```

## Creating Derived Variables – Symptom Status Flag

I create a clean binary flag `has_no_symptoms` (TRUE/FALSE) to identify participants who reported having **no symptoms** at all.  

```{r has-no-symptoms-flag}
# Create the binary flag
yao_symptoms <- yaounde_covid %>%
  mutate(
    has_no_symptoms = symptoms == "No symptoms"
  ) %>%
  select(symptoms, has_no_symptoms)

# Interactive preview with reactable – portfolio-friendly
reactable(
  yao_symptoms,
  defaultPageSize = 10,
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  bordered = TRUE,
  showPagination = TRUE,            # this is key
  searchable = TRUE,                    # allows quick filtering/search in browser
  columns = list(
    symptoms = colDef(
      name = "Reported Symptoms",
      minWidth = 220
    ),
    has_no_symptoms = colDef(
      name = "No Symptoms?",
      align = "center"
    )
))
```

## Sarcopenia Dataset – Creating Grip Strength Indicator

In the sarcopenia dataset for elderly participants, grip strength is a key diagnostic criterion.  

Low grip strength is often defined with sex-specific cut-offs, but for illustration here we use a simple threshold of < 20 (common approximate value for identifying weakness in older adults, though clinical cut-offs vary by guideline and population).

We create a binary flag `low_grip_strength` (TRUE = grip strength < 20) to support later analysis (e.g. prevalence of low grip strength, association with other variables).

```{r grip-strength-flag}
# Create the binary indicator
sacro_grip <- sacro_data %>%
  select(grip_strength_kg) %>%
  mutate(
    low_grip_strength = grip_strength_kg < 20
  )

# Summary table – shows distribution of the new flag
sacro_grip %>%
  count(low_grip_strength) %>%
  mutate(
    status  = if_else(low_grip_strength, "Low grip strength (< 20)", "Normal / Higher grip strength"),
    percent = round(n / sum(n) * 100, 1)
  ) %>%
  select(status, n, percent) %>%
  gt::gt() %>%
  tab_header(
    title = "Grip Strength Classification (Threshold < 20)",
    subtitle = paste("n =", nrow(sacro_grip), "participants with grip strength recorded")
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  ) %>%
  cols_label(
    status  = "Grip Strength Status",
    n       = "Number of Participants",
    percent = "Percentage (%)"
  ) %>%
  fmt_number(
    columns = percent,
    decimals = 1
  ) %>%
  # Bold column headers
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  # Optional: subtle row highlighting for low grip
  tab_style(
    style = cell_fill(color = "#fff5f5"),
    locations = cells_body(
      columns = everything(),
      rows = status == "Low grip strength (< 20)"
    )
  )

```

## Age Group Classification – Influenza Dataset

We categorize participants into "Child" (< 18 years) and "Adult" (≥ 18 years) using the `age` variable.  

Missing age values are explicitly labeled as "Missing Age" to make downstream analysis transparent and avoid silent dropping of records.

```{r}
influenza %>% 
  mutate(age_group = case_when(age < 18 ~ "Child"
                               , age >= 18 ~ "Adult",
                               is.na(age) ~ "Missing Age")) %>%
  select(age, age_group) %>%
  count(age_group) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1)
  ) %>%
  arrange(match(age_group, c("Child", "Adult", "Missing Age"))) %>%
  gt::gt() %>%
  tab_header(
    title = "Age Group Distribution"
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  ) %>%
  cols_label(
    age_group = "Age Group",
    n         = "Count",
    percent   = "Percentage (%)"
  ) %>%
  fmt_number(
    columns = percent,
    decimals = 1
  ) %>%
  # Bold headers
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  # Subtle highlighting for visual distinction
  tab_style(
    style = cell_fill(color = "#f0f8ff"),
    locations = cells_body(rows = age_group == "Child")
  ) %>%
  tab_style(
    style = cell_fill(color = "#fff5e6"),
    locations = cells_body(rows = age_group == "Adult")
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f8f8"),
    locations = cells_body(rows = age_group == "Missing Age")
  )
  
```

## BMI Calculation and Classification – Yaoundé COVID-19 Dataset

We calculate **Body Mass Index (BMI)** using the standard formula:

BMI = weight (kg) / [height (m)]²

Height is converted from cm to meters, and BMI is rounded to 1 decimal place for readability.  

Participants are then classified into standard WHO BMI categories:

- Underweight: < 18.5

- Normal weight: 18.5–24.9

- Overweight: 25–29.9

- Obese: ≥ 30

(Missing or implausible height/weight values will result in NA BMI values.)

```{r calculate-bmi-and-classify}
yaounde_bmi <- yaounde_covid %>%
  mutate(
    height_m   = height_cm / 100,
    bmi        = round(weight_kg / (height_m ^ 2), 1),
    bmi_class  = case_when(
      is.na(bmi)                          ~ "Missing / Invalid data",
      bmi < 18.5                          ~ "Underweight",
      bmi >= 18.5 & bmi < 25              ~ "Normal weight",
      bmi >= 25   & bmi <= 29.9           ~ "Overweight",     # <= 29.9 to avoid overlap
      bmi > 29.9                          ~ "Obese"    )
  ) %>%
  select(height_cm, weight_kg, height_m, bmi, bmi_class) 

yaounde_bmi %>% 
  count(bmi_class) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1)
  ) %>%
  arrange(match(bmi_class, c("Underweight", "Normal weight", "Overweight", "Obese", "Missing / Invalid data"))) %>%
  gt::gt() %>%
  tab_header(
    title = "BMI Category Distribution",
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(14)
  ) %>%
  cols_label(
    bmi_class = "BMI Category",
    n         = "Number of Participants",
    percent   = "Percentage (%)"
  )
```



----
Proverbs 11:28
